Forecasting the 2024 US Presidential Election

### 2024-11-04 paper/paper.qmd
Finished discussion section, added a reference to the reference bib.tex file, and rendered the pdf.

## Evaluating the Model

@@

ggplot(data = final_pooled_both, aes(x = Week)) + 
  geom_line(aes(y = Harris, colour = "Harris")) + 
  geom_line(aes(y = Trump, colour = "Trump")) +
  geom_point(aes(y = Harris), color = "blue") +
  geom_point(aes(y = Trump), color="red") +
  theme_light() +
  labs(
    y = "Percentage of Support (%)",
    caption = "Data Source: 538"
  ) + 
  ylim(40,55) + scale_x_continuous(name = "Week", breaks = seq(1,12,by=1), limits = c(1, 12)) + 
  ggtitle("Percentage of Support For Presidential Candidates Over Time") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(name = "Candidate", values = c("Harris" = "blue", "Trump" = "red"))

@@

ggplot(pooled_and_predicted_harris, aes(Week, Predictedpct)) +
  geom_point(colour = "blue") +
  geom_smooth(method='lm', color='black') +
  theme_light() +
  labs(x = "Week", y = "Percentage of Support (pct)", 
       title = "Predicted Percentage of Support (pct) for Harris Over Time") +
  theme(plot.title = element_text(hjust=0.5)) +
  scale_x_continuous(name = "Week", breaks = seq(1,12,by=1), limits = c(1, 12)) +
  geom_line(aes(y = WeightedAveragepct, colour = "Weighted Average pct")) +
  scale_color_manual(name = "Data", values = c("Weighted Average pct" = "darkgreen"))

@@

ggplot(pooled_and_predicted_trump, aes(Week, Predictedpct)) +
  geom_point(colour = "red") +
  geom_smooth(method='lm', color='black') +
  theme_light() +
  labs(x = "Week", y = "Percentage of Support (pct)", 
       title = "Predicted Percentage of Support (pct) for Trump Over Time") +
  theme(plot.title = element_text(hjust=0.5)) +
  scale_x_continuous(name = "Week", breaks = seq(1,12,by=1), limits = c(1, 12)) +
  geom_line(aes(y = WeightedAveragepct, colour = "Weighted Average pct")) +
  scale_color_manual(name = "Data", values = c("Weighted Average pct" = "darkgreen"))

@@

ggplot(data = final_pooled_both, aes(x = Week)) + 
  geom_line(aes(y = Harris, colour = "Harris")) + 
  geom_line(aes(y = Trump, colour = "Trump")) +
  geom_point(aes(y = Harris), color = "blue") +
  geom_point(aes(y = Trump), color="red") +
  theme_light() +
  labs(
    y = "Percentage of Support (%)",
    caption = "Data Source: 538"
  ) + 
  ylim(46,50) + scale_x_continuous(name = "Week", breaks = seq(1,13,by=1), limits = c(1, 13)) + 
  ggtitle("Forecasted Percentages of Support for Harris and Trump in Week 13") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(name = "Candidate", values = c("Harris" = "blue", "Trump" = "red")) +
  geom_point(aes(x = 13, y = 48.03), colour = "blue", size = 1.5) + 
  geom_point(aes(x = 13, y = 47.98), colour = "red", size = 1.5)

@@

## Summary

*A paragraph that summarizes what was done in the analysis and a brief overview of the main findings*

## Implications

*A paragraph about what the main findings imply about the election - why they are relevant*

## Limitations

*A paragraph about the limitations of the model - shortcomings due to the decisions that were made in the model*

## Future Directions

*A paragraph about what can be done in the future*

>>>>

## Evaluating the Model {#sec-evaluation}

@@

ggplot(data = final_pooled_both, aes(x = Week)) + 
  geom_line(aes(y = Harris, colour = "Harris")) + 
  geom_line(aes(y = Trump, colour = "Trump")) +
  geom_point(aes(y = Harris), color = "blue") +
  geom_point(aes(y = Trump), color="red") +
  theme_light() +
  labs(
    y = "Percentage of Support (%)",
    caption = "Data Source: 538"
  ) + 
  ylim(40,55) + scale_x_continuous(name = "Week", breaks = seq(1,12,by=1), limits = c(1, 12)) + 
  ggtitle("Percentage of Support For Presidential Candidates Over Time") +
  theme(plot.title = element_text(hjust = 0.2)) +
  scale_color_manual(name = "Candidate", values = c("Harris" = "blue", "Trump" = "red"))

@@

ggplot(pooled_and_predicted_harris, aes(Week, Predictedpct)) +
  geom_point(colour = "blue") +
  geom_smooth(method='lm', color='black') +
  theme_light() +
  labs(x = "Week", y = "Percentage of Support (pct)", 
       title = "Predicted Percentage of Support (pct) for Harris Over Time") +
  theme(plot.title = element_text(hjust=0.2)) +
  scale_x_continuous(name = "Week", breaks = seq(1,12,by=1), limits = c(1, 12)) +
  geom_line(aes(y = WeightedAveragepct, colour = "Weighted Average pct")) +
  scale_color_manual(name = "Data", values = c("Weighted Average pct" = "darkgreen"))

@@

ggplot(pooled_and_predicted_trump, aes(Week, Predictedpct)) +
  geom_point(colour = "red") +
  geom_smooth(method='lm', color='black') +
  theme_light() +
  labs(x = "Week", y = "Percentage of Support (pct)", 
       title = "Predicted Percentage of Support (pct) for Trump Over Time") +
  theme(plot.title = element_text(hjust=0.2)) +
  scale_x_continuous(name = "Week", breaks = seq(1,12,by=1), limits = c(1, 12)) +
  geom_line(aes(y = WeightedAveragepct, colour = "Weighted Average pct")) +
  scale_color_manual(name = "Data", values = c("Weighted Average pct" = "darkgreen"))

```
\newpage

@@

ggplot(data = final_pooled_both, aes(x = Week)) + 
  geom_line(aes(y = Harris, colour = "Harris")) + 
  geom_line(aes(y = Trump, colour = "Trump")) +
  geom_point(aes(y = Harris), color = "blue") +
  geom_point(aes(y = Trump), color="red") +
  theme_light() +
  labs(
    y = "Percentage of Support (%)",
    caption = "Data Source: 538"
  ) + 
  ylim(46,50) + scale_x_continuous(name = "Week", breaks = seq(1,13,by=1), limits = c(1, 13)) + 
  ggtitle("Forecasted Percentages of Support in Week 13") +
  theme(plot.title = element_text(hjust = 0.1)) +
  scale_color_manual(name = "Candidate", values = c("Harris" = "blue", "Trump" = "red")) +
  geom_point(aes(x = 13, y = 48.03), colour = "blue", size = 1.5) + 
  geom_point(aes(x = 13, y = 47.98), colour = "red", size = 1.5)

@@

As the 2024 US election approaches, the present analysis forecasts the percentage of support both presidential candidates will receive in the final week leading up to election day on Tuesday November 5, 2024. Through the use of seasonal indexes and a regression model, this analysis takes in to account how voter opinions or preferences change over time to generate more precise forecasts. With this, an understanding of how support for Harris and Trump has changed over time is provided in addition to the model’s forecasts.

This analysis finds that while support for both Harris and Trump fluctuate over the course of the 12 weeks defined by the model, overall support for Harris appears to increase over time. In contrast, despite having a high percentage of support in Week 1, Trump is found to have received less support in Week 12. This pattern indicates that the percentages of support for Harris and Trump are converging as time goes on, implying that the election will be a tight race that may need to be determined by the swing states. Additionally, the forecasts computed by the model show that Harris will receive greater support than Trump during the week before the election (Week 13). However, as the difference in support that Harris and Trump will receive during this time is extremely small (0.05%), this result also gives rise to the notion that the 2024 election will continue to be a close race until election day.

As briefly mentioned in @sec-evaluation, a limitation of this analysis is that it assumes that polls are unbiased to pool them together. By not considering the effects of methodology, vote population (e.g. registered voters vs. likely voter), different states, and poll quality, the model may be amplifying the biases of different polls as it pools them together [@journal]. Another limitation of this model is that it considers 12 weeks of data to forecast. Though this is a sufficient amount of data for the model, a stronger representation of the change in candidate support over time as well as more robust forecasts may be produced with a larger quantity of data.

Given these limitations, moving forward, this analysis should be adapted to incorporate a way to account for variation across polls. With a model that considers the effects of methodology, vote population, states, and poll quality along with seasonal indexes, more precise and less biased estimates of the percentage of support for candidates during a given time period could be made. As the findings of this analysis point to a very close race that can be determined by the swing states, a next step for the model could be to forecast the percentage of support for each candidate within each swing state to present a more detailed prediction of the election. Lastly, it may also be helpful to adopt a more sophisticated method to pool the polls, such as pooling the polls by considering the margin of error of each poll [@journal]. A more enhanced method may allow for better precision, and in turn, provide a more reliable way to forecast.
